\bf Approximating Likelihood Ratios with Calibrated Discriminative Classifiers

### 2016-03-13 jasa/paper-jasa.tex
cleaning up example 1 text

%    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc.
%      %(or a blank line to force the subfigure onto a new line)
%    \begin{subfigure}[b]{0.3\textwidth}
%        \includegraphics[width=\textwidth]{figures/fig_ratio_dist.png}
%        \caption{Likelihood ratio distributions}
%        \label{fig:decomposed_ratio}
%    \end{subfigure}
    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc.
      %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[width=\textwidth]{figures/fig-ratio-dist.pdf}
        \caption{$p( \log \hat r(\mathbf{x})) \, | \,\gamma = 0.05)$ }
        \label{fig:decomposed_ratio}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[width=\textwidth]{figures/fig-likelihood.pdf}
        \caption{log-likelihood curve }
        \label{fig:likelihood}
    \end{subfigure}

@@

While not directly apparent, calibrating $\hat s(\mathbf{x})$ to construct $\hat p(\hat
s(\mathbf{x}))$ is key to obtain accurate results. Standard
histograms with fixed bins have been used here for illustrative purposes, but we
anticipate that the embedded density estimation algorithm should be carefully
chosen and tuned, on a case by case basis. Overall, this first example confirms
that even if the classifier does a poor job at modeling the optimal decision
function, good approximations of the likelihood ratio can still be obtained
with the proposed method, by calibrating $s(\mathbf{x})$ and/or by decomposing
the mixture when possible.
%
%
%\begin{figure}
%    \centering
%    \begin{subfigure}[b]{0.3\textwidth}
%        \includegraphics[width=\textwidth]{figures/p_sij.png}
%        \caption{Pairwise classifier score distributions}
%        \label{fig:p_sij}
%    \end{subfigure}
%    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc.
%      %(or a blank line to force the subfigure onto a new line)
%    \begin{subfigure}[b]{0.3\textwidth}
%        \includegraphics[width=\textwidth]{figures/fig_ratio_dist.png}
%        \caption{Likelihood ratio distributions}
%        \label{fig:decomposed_ratio}
%    \end{subfigure}
%    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc.
%      %(or a blank line to force the subfigure onto a new line)
%    \begin{subfigure}[b]{0.3\textwidth}
%        \includegraphics[width=\textwidth]{figures/likelihood.png}
%        \caption{log-likelihood curve }
%        \label{fig:likelihood}
%    \end{subfigure}
%
%    \caption{Approximation of the log-likelihood ratio
%             $\log r(\mathbf{x};\gamma=0.05,\gamma=0) = \log \frac{p(\mathbf{x}|\gamma=0.05)}{p(\mathbf{x}|\gamma=0)}$
%             with calibrated classifiers.}
%\end{figure}
%
%%%%%%%%%%%%%

@@

Given observed data from Figure~\ref{fig:1a}, let us now consider the composite hypothesis testing setting, by comparing
$H_0: \gamma = 0.05$ versus the alternative $H_1: \gamma \neq 0.05$. That is, we want to evaluate
\begin{equation}
\Lambda( \gamma = 0.05) =  \frac{ p({\cal D} | \gamma = 0.05) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} \; .
\end{equation}
%\begin{equation}
%\Lambda({\cal D}; \Theta_0 = \{0.05\}, \Theta=[0, 1]) =  \frac{ p({\cal D} | \gamma = 0.05) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} \; .
%\end{equation}
By exploiting the fact that
\begin{equation}
\log \frac{ p({\cal D} | \gamma = 0.05) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} = \log \frac{ p({\cal D} | \gamma = 0.05) }{ p({\cal D} | \gamma = 0 ) } - \log \frac{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma) }{  p({\cal D} | \gamma = 0 ) },
\end{equation}
the generalized likelihood ratio statistic can be computed by evaluating
both terms reusing the decomposed neural network-based approximation from Figure~\ref{fig:1c}.
In particular, the maximum likelihood estimator (MLE) in the second term
can be found as outlined in
Section~\ref{sec:generalized-likelihood-ratio}.
Figure~\ref{fig:2a} shows the empirical distribution of the resulting MLEs,
compared to the distribution of the true MLEs. It clearly suggests that
the proposed inference procedure yields in this case an unbiased estimator
($\mathbb{E} [\hat \gamma] = 0.0501$, as empirically estimated over $1000$
trials). Note that had the underlying ratio not been accurate enough because of
poor training or poor calibration, the two distributions would not be exactly
overlapping, indicating bias. From this, the initial $\Lambda$ test statistic can finally
be evaluated to determine in this case that the null should not be rejected.  Figure \ref{fig:2b} also confirms
that $-2 \log \Lambda$, as approximated with our method, is asymptotically $\chi^2_1$ distributed,
as should be the true statistic when the null is true.

>>>>

%    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc.
%      %(or a blank line to force the subfigure onto a new line)
%    \begin{subfigure}[b]{0.3\textwidth}
%        \includegraphics[width=\textwidth]{figures/fig_ratio_dist.png}
%        \caption{Likelihood ratio distributions}
%        \label{fig:decomposed_ratio}
%    \end{subfigure}
    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc.
      %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[width=\textwidth]{figures/fig-ratio-dist.pdf}
        \caption{$p( \log \hat r(\mathbf{x})) \, | \,\gamma = 0.05)$ }
        \label{fig:decomposed_ratio}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[width=\textwidth]{figures/fig-likelihood.pdf}
        \caption{$-\log \Lambda(\gamma)$ }
        \label{fig:likelihood}
    \end{subfigure}

@@

As the results show, calibrating $\hat s(\mathbf{x})$ through univariate density estimation of 
$\hat p(\hat s(\mathbf{x}))$ is key to obtaining accurate results. Standard
histograms with uniform binning have been used here for illustrative purposes, but we
anticipate that more sophisticated calibration strategies will be important in further development
of this method. We leave this as an area for future study.

Figure~\ref{fig:decomposed_ratio} shows the distribution of $ \log \hat r(\mathbf{x}))$ for $\gamma=0.05$
using the decomposed approximation of $\hat r(\mathbf{x}))$ with neural networks and the distribution
of the exact log-likelihood ratio. While there are some artifacts in the distribution in the low-probability regions and the maximum value of the log-likehood ratio is underestimated, the overall shape of the distribution is well approximated.

Finally, we see the log-likelihood curve $-\log \Lambda(\gamma)$ for the dataset $\mathcal{D}$ shown in 
Fig.~\ref{fig:1a}. The exact likelihood curve is very well approximated by the method, confirming that
even when the raw classifier does a poor job at modeling the $s^*(\mathbf{x}$, a good approximations of the 
likelihood ratio can still be obtained by calibrating $s(\mathbf{x})$ (and by decomposing
the mixture, if possible).

Given observed data from Figure~\ref{fig:1a}, let us now consider the composite hypothesis testing setting, by comparing
$H_0: \gamma = 0.05$ versus the alternative $H_1: \gamma \neq 0.05$. That is, we want to evaluate
\begin{equation}
\Lambda( \gamma = 0.05) =  \frac{ p({\cal D} | \gamma = 0.05) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} \; .
\end{equation}
%\begin{equation}
%\Lambda({\cal D}; \Theta_0 = \{0.05\}, \Theta=[0, 1]) =  \frac{ p({\cal D} | \gamma = 0.05) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} \; .
%\end{equation}
By exploiting the fact that
\begin{equation}
\log \frac{ p({\cal D} | \gamma = 0.05) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} = \log \frac{ p({\cal D} | \gamma = 0.05) }{ p({\cal D} | \gamma = 0 ) } - \log \frac{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma) }{  p({\cal D} | \gamma = 0 ) },
\end{equation}
the generalized likelihood ratio statistic can be computed by evaluating
both terms reusing the decomposed neural network-based approximation from Figure~\ref{fig:1c}.
In particular, the maximum likelihood estimator (MLE) in the second term
can be found as outlined in
Section~\ref{sec:generalized-likelihood-ratio}.

@@

Figure~\ref{fig:2a} shows the empirical distribution of the resulting MLEs,
compared to the distribution of the true MLEs. It clearly suggests that
the proposed inference procedure yields in this case an unbiased estimator
($\mathbb{E} [\hat \gamma] = 0.0501$, as empirically estimated over $1000$
trials). Note that had the underlying ratio not been accurate enough because of
poor training or poor calibration, the two distributions would not be exactly
overlapping, indicating bias. From this, the initial $\Lambda$ test statistic can finally
be evaluated to determine in this case that the null should not be rejected.  Figure \ref{fig:2b} also confirms
that $-2 \log \Lambda$, as approximated with our method, is asymptotically $\chi^2_1$ distributed,
as should be the true statistic when the null is true.
